The R\"{o}ssler System

### 2016-11-18 ode_two/paper.tex
finished prezzie

R\"{o}ssler's analysis focused on the behavior of the system for small $a$ and $b$. For simplicity, we can consider the solutions to (\ref{eq:csln1}) and (\ref{eq:csln2}) for very small $a$ and $b$. We now have
\begin{gather}
	\left[
	x=0,
	y=0,
	z=0
	\right]\label{eq:csln1sm}\\
	\left[
	x=c,
	y=-\frac{c}{a},
	z=\frac{c}{a}
	\right]\label{eq:csln2sm}.
\end{gather}
To find the behavior around the critical points given by (\ref{eq:csln1sm}), we evaluate the Jacobian at the appropriate points and find the eigenvalues of the system.
\[
	|J(0,0,0)-\lambda I| =
	\left|
	\left[
	\begin{array}{c c c}
		-\lambda & -1       & -1         \\
		1        & -\lambda & 0          \\
		0        & 0        & -c-\lambda
	\end{array}
	\right]
	\right| =
	(c-\lambda)(\lambda + 1)^{2}=0.
\]
Hence, the system has three eigenvalues: $\lambda=-i, \lambda=-i,$ and $\lambda = -c$ (we will assume here that $c > 1$). This implies that the system has an unstable focus node at the origin \cite{equilibria}. Figure \ref{fig:plot_cuts} shows various cuts of the system for various $z$ values; the system spirals in the $x-y$ plane, while moving towards and then away from the origin as $z$ moves from $-\infty$ to $\infty$. We can solve this linearized system by looking at the equation
\[
	\mathbf{\dot{x}} =
	\left[
	\begin{array}{c c c}
		0 & -1 & -1 \\
		1 & 0  & 0  \\
		0 & 0  & -c
	\end{array}
	\right]
	\mathbf{x}.
\]
Since we've already found the eigenvalues of the system, we can find the eigenvectors immediately.

\begin{minipage}{0.3\textwidth}
	\begin{gather*}
		\text{\underline{$\lambda=-i$}}\\
		\mathbf{\lambda_{1}}=\left[
		\begin{array}{c}
			-i \\
			1  \\
			0
		\end{array}
		\right].
	\end{gather*}
\end{minipage}
\begin{minipage}{0.3\textwidth}
	\begin{gather*}
		\text{\underline{$\lambda=i$}}\\
		\mathbf{\lambda_{2}}=\left[
		\begin{array}{c}
			i \\
			1 \\
			0
		\end{array}
		\right].
	\end{gather*}
\end{minipage}
\begin{minipage}{0.3\textwidth}
	\begin{gather*}
		\text{\underline{$\lambda=-c$}}\\
		\mathbf{\lambda_{3}}=\left[
		\begin{array}{c}
			\frac{c}{1+c^{2}}  \\
			-\frac{1}{1+c^{2}} \\
			1
		\end{array}
		\right].
	\end{gather*}
\end{minipage}

\noindent\\
\noindent Letting $\mathbf{\gamma}$ stand in for our constants, we get
\[
	\mathbf{x}(t) =
	\gamma_{1}e^{-it}\left[
	\begin{array}{c}
		-i \\
		1  \\
		0
	\end{array}
	\right] +
	\gamma_{2}e^{it}
	\left[
	\begin{array}{c}
		i \\
		1 \\
		0
	\end{array}
	\right] +
	\gamma_{3}e^{-ct}\left[
	\begin{array}{c}
		\frac{c}{1+c^{2}}  \\
		-\frac{1}{1+c^{2}} \\
		1
	\end{array}
	\right].
\]
Note that we choose to leave the $e^{it}$ in this form, since \textbf{${\lambda_{1}}$} and $\mathbf{\lambda_{2}}$ are linearly independent of each other. Figure \ref{fig:zero_sln} shows the behavior of this linearized system with two different sets of initial conditions. If we compare this to figure \ref{fig:3dsys_02}, we see similar behavior in the area of the graph with large $z$ values.

>>>>

R\"{o}ssler's analysis focused on the behavior of the system for small $a$ and $b$. For simplicity, we can consider the solutions to (\ref{eq:csln1}) and (\ref{eq:csln2}) for very small $a$ and $b$. We now have
\begin{gather}
	\left[
	x=0,
	y=0,
	z=0
	\right]\label{eq:csln1sm}\\
	\left[
	x=c,
	y=-\frac{c}{a},
	z=\frac{c}{a}
	\right]\label{eq:csln2sm}.
\end{gather}
To find the behavior around the critical points given by (\ref{eq:csln1sm}), we evaluate the Jacobian at the appropriate points and find the eigenvalues of the system.
\[
	|J(0,0,0)-\lambda I| =
	\left|
	\left[
	\begin{array}{c c c}
		-\lambda & -1       & -1         \\
		1        & -\lambda & 0          \\
		0        & 0        & -c-\lambda
	\end{array}
	\right]
	\right| =
	(-c-\lambda)(\lambda^{2} + 1)=0.
\]
Hence, the system has three eigenvalues: $\lambda=-i,\ \lambda=i,$ and $\lambda = -c$ (we will assume here that $c > 1$). This implies that the linear system has a center at the origin \cite{equilibria}. Since centers are \textbf{not} hyperbolic, we can look at various stream plots of the system. Figure \ref{fig:plot_cuts} shows various cuts of the system for various $z$ values; the system spirals in the $x-y$ plane, while moving towards and then away from the origin as $z$ moves from $-\infty$ to $\infty$. Hence, the system appears to actually have a spiral at the origin. We can solve this linearized system by looking at the equation
\[
	\mathbf{\dot{x}} =
	\left[
	\begin{array}{c c c}
		0 & -1 & -1 \\
		1 & 0  & 0  \\
		0 & 0  & -c
	\end{array}
	\right]
	\mathbf{x}.
\]
Since we've already found the eigenvalues of the system, we can find the eigenvectors immediately.

\begin{minipage}{0.3\textwidth}
	\begin{gather*}
		\text{\underline{$\lambda=-i$}}\\
		\mathbf{\lambda_{1}}=\left[
		\begin{array}{c}
			-i \\
			1  \\
			0
		\end{array}
		\right].
	\end{gather*}
\end{minipage}
\begin{minipage}{0.3\textwidth}
	\begin{gather*}
		\text{\underline{$\lambda=i$}}\\
		\mathbf{\lambda_{2}}=\left[
		\begin{array}{c}
			-i \\
			1 \\
			0
		\end{array}
		\right].
	\end{gather*}
\end{minipage}
\begin{minipage}{0.3\textwidth}
	\begin{gather*}
		\text{\underline{$\lambda=-c$}}\\
		\mathbf{\lambda_{3}}=\left[
		\begin{array}{c}
			\frac{c}{1+c^{2}}  \\
			-\frac{1}{1+c^{2}} \\
			1
		\end{array}
		\right].
	\end{gather*}
\end{minipage}

\noindent\\
\noindent Letting $\mathbf{\gamma}$ stand in for our constants, we get
\[
	\mathbf{x}(t) =
	\gamma_{1}e^{-it}\left[
	\begin{array}{c}
		-1 \\
		1  \\
		0
	\end{array}
	\right] +
	\gamma_{2}e^{it}
	\left[
	\begin{array}{c}
		-1 \\
		1 \\
		0
	\end{array}
	\right] +
	\gamma_{3}e^{-ct}\left[
	\begin{array}{c}
		\frac{c}{1+c^{2}}  \\
		-\frac{1}{1+c^{2}} \\
		1
	\end{array}
	\right].
\]
Note that we choose to leave the $e^{it}$ in this form, since \textbf{${\lambda_{1}}$} and $\mathbf{\lambda_{2}}$ are linearly independent of each other. Figure \ref{fig:zero_sln} shows the behavior of this linearized system with two different sets of initial conditions. If we compare this to figure \ref{fig:3dsys_02}, we see similar behavior in the area of the graph with large $z$ values.

### 2016-11-18 ode_two/presentation.tex
finished prezzie

\begin{frame}{Linearization}
    \begin{itemize}
        \item Jacobian of system will tell us about behavior of fixed points
    \end{itemize}

\end{frame}

>>>>

\begin{frame}{Linearization}
    \begin{itemize}
        \item Jacobian of system will tell us about behavior of fixed points
		\pause
    \end{itemize}
	\[
		|J(0,0,0)-\lambda I| =
		\left|
		\left[
		\begin{array}{c c c}
			-\lambda & -1       & -1         \\
			1        & -\lambda & 0          \\
			0        & 0        & -c-\lambda
		\end{array}
		\right]
		\right| =
		(c-\lambda)(\lambda + 1)^{2}=0.
	\]
	\pause
	\begin{itemize}
		\item System has three eigenvalues
		\pause
	\end{itemize}
	\[
		\lambda=-i\qquad\pause \lambda=-i\qquad\pause\lambda = -c
	\]
	\pause
	\begin{definition}
		If two of the eigenvalues of the system are pure imaginary, then the system has a \alert{center} at the fixed point. This is \textbf{not} a hyperbolic equilibrium.
    \end{definition}
	\pause
	\begin{itemize}
		\item $\therefore (0,0,0)$ is a center of the linear system
	\end{itemize}
\end{frame}

\begin{frame}{Analyzing the Center}
	\begin{itemize}
		\item How to analyze the center of linear system?
		\pause
		\item Hold $z$ constant
		\pause
		\item Look at behavior of $(0,0)$ in $x-y$ plane
	\end{itemize}
	\begin{center}
		\begin{tabular}{c c}
			\includegraphics[scale=0.3]{spz-2} & \includegraphics[scale=0.3]{spz2}\\
			$z=-2$ & $z=2$
		\end{tabular}
	\end{center}
	\pause
	\begin{itemize}
		\item Actually appears to be a spiral
	\end{itemize}
\end{frame}

\begin{frame}{Linearization}
	\begin{itemize}
		\item Can analyze the other fixed point
		\pause
	\end{itemize}
	\begin{gather*}
		\left|J\left(c,-\frac{c}{a},\frac{c}{a}\right) -\lambda I\right| =
		\frac{ ac - a\lambda - c\lambda + a^{2}\lambda^{2} -a\lambda^{3}}{a}.
	\end{gather*}
	\pause
	\begin{itemize}
		\item Solving with Mathematica yields bad values
		\pause
		\item Recall system was meant to be analyzed with small $a,b$
		\pause
		\item Let $a\to0.14$, $c\to8.8$
	\end{itemize}
	\pause
	\[
		\lambda =0.001-7.99i\pause\qquad \lambda=0.001+7.99i\pause\qquad\lambda=0.13
	\]
	\pause
	\begin{itemize}
		\item $\therefore$ fixed point is an unstable focus node
	\end{itemize}
\end{frame}

\begin{frame}{Chaos}
	\begin{center}
		\includegraphics[scale=0.2]{bane}
	\end{center}
	\pause
	\begin{itemize}
		\item Mathematica demonstration
	\end{itemize}
\end{frame}

\begin{frame}{Chaos}
	\begin{itemize}
		\item Small changes in the system yield drastic changes in behavior
		\pause
		\item Periodicity changes to chaotic behavior
	\end{itemize}
	\pause
	\begin{center}
		\includegraphics[scale=0.3]{varying_c}
	\end{center}
\end{frame}

%can talk about atrractiveness if need more time
\begin{frame}{References}
	\nocite{*}
	\bibliographystyle{alpha}
	\bibliography{mbib}
\end{frame}
